Spatial Reasoning and Statistical Graphics

### 2014-10-20 Paper/Draft1/TestingVisualAptitude.Rnw
Lit review update

\section{Introduction}
Relevant literature: 
\begin{itemize}
\item \citet{shah1995conceptual} showed that spatial ability was not correlated with accuracy on a simple two-dimensional line graph description task, but that mathematical ability was correlated with accuracy. 
\item \citet{just1985cognitive} showed that high-spatial-ability viewers used different rotation strategies than low-spatial-ability viewers when asked to whether three-dimensional alphabet cubes were the same. 
\item \citet{hofmann2012graphical} for lineup stimuli and general lineup performance
\end{itemize}

>>>>

\section{Introduction}
Relevant literature: 
\item \citet{shah1995conceptual} showed that spatial ability was not correlated with accuracy on a simple two-dimensional line graph description task, but that mathematical ability was correlated with accuracy. 
\item \citet{just1985cognitive} showed that high-spatial-ability viewers used different rotation strategies than low-spatial-ability viewers when asked to whether three-dimensional alphabet cubes were the same. 
\item \citet{voyer1995magnitude} completed a meta-analysis of spatial tasks and gender differences. They concluded that ``rotation" tasks, such as the card rotation test, have robust sex differences across cohort, and while the sex differences seem to be declining for some tests, such as the card rotation test, the differences still exist for the time being. Test administration differences may account for the change. They also concluded that the paper-folding test showed no such sex difference, though this may be dependent on test scoring (a lower guessing penalty decreases the score difference). 
\comment{\textit{In the current analysis, we have scored the paper folding test out of 20. Alternate scoring procedures still need to be investigated... they cite a poster, but it's not available anywhere online as far as I can tell.}}
\item \citet{lowrie2007solving} investigated students (9-10 yrs) ability to read mathematical graphs and their visual-spatial ability. They partition graphs into 6 ``graphical languages" (some of these languages aren't statistical graphs - networks, venn diagrams, maps), and use multivariate analysis to associate spatial ability with some of the types of graphs studied. 
\comment{\textit{I can't understand their model results very well, so I'm not entirely sure what's going on here. \\
They also cite a bunch of studies (including \citet{voyer1995magnitude}) that I haven't had time to completely explore.}}
\item \citet{vekiri2002value} is lit review discussing the different theories about the utility of graphics in learning (dual coding theory, visual argument hypothesis, and conjoint retention hypothesis). The article contains a nice definition of graphics (though it might be too domain specific for statistical graphics, since it claims graphics have only one meaning and aren't prone to interpretation.) Depending on the context, ``maps", ``charts", and ``graphs" here could fall into a loose classification as ``statistical graphics", but in general ``graphs" is probably accurate for most of the lineups we're testing: ``[Referents are ] quantitative data ... that enable viewers to compare and observe relations among variables". 
\item \textbf{Dual Coding Theory}: people encode both words and pictures separately, so both are encoded for a graph. \\
Supports the use of graphics, and there is support for dual-coding in working memory research (pictures and words are processed separately but in parallel). In addition, since graphics depend on other abilities (spatial abilities), graphics that aren't properly designed create higher cognitive load and lower performance on graphical tasks. The review also specifically discusses visuospatial ability, citing a speculation from \citet{mayer1994whom} that students with low spatial ability perform worse with diagrams and graphs because more cognitive resources are required.
\comment{\textit{Lots of studies by Sweller cited throught the cognitive load discussion - I haven't gotten to that literature yet.\\That said, that research has some potential to justify the use of lineups from a psychological perspective - if you can claim that types of graphs that have poor performance do so because they create higher cognitive load (cue psych research into why that would be true, cleveland \& mcgill, etc.), you can justify the suckiness of those awful boxplot-jittered-things.}}
\item \textbf{Visual Argument Hypothesis}: Graphics require less processing because their visuospatial properties encode some of the information through the arrangement of elements as well as the elements themselves. That is, graphics are a form of ``external cognition" \citep{scaife1996external} that guide, constrain, and facilitate cognitive behavior\citep{zhang1997nature}. This constraint reduces memory load and makes more cognitive resources available for other tasks; displays that clearly present information without deep processing requirements are more effective\citep{zhang1994representations}. Graphical displays have higher search and computational efficiency than text \citep{larkin1987diagram}, and when graphics are designed according to gestalt principles of organization, where grouping lines up with visual chunks, gestalt heuristics can be used to more quickly encode information \citep{shah1999graphs}.
\comment{\textit{I am working on making my way through all of this literature - right now, I'm re-citing stuff; I will print the papers out and add to this ASAP once I get toner into the printer.}}
\item \textbf{Conjoint Retention Hypothesis} - based on dual coding theory, but claims that there are separate but interconnected memory codes for verbal and visual information. Assumes maps/graphs are encoded intact, with visuospatial properties preserved. This is pretty unlikely, and the literature is not so relevant to our cause. Spatial information is encoded, but words can actually interfere with this encoding; it is not ``intact" 
\item \citet{mayer1994whom} used the card rotation test (10 figures, 80 questions total) and the paper folding test (10 items) to evaluate spatial performance, along with a problem solving test. They used the same penalty structure on the card rotation test, and don't seem to have used any penalty on the paper folding test; they then scaled the two tests. Their problem solving test was more practical (engineering-based), so isn't applicable to our study. They were testing instruction using visual + audio in sync, visual + audio out of sync, and then no instruction as a control. High-ability students could make more use of the in-sync presentation of visual and auditory information than low-ability students, but did not differ significantly in the out-of-sync information condition. Thus, visualization ability moderates the results of the information presented by allowing students to process information concurrently more efficiently, but doesn't help students process the information separately or help students come up with solutions to the problem-solving task. 
\item \citet{hofmann2012graphical} for lineup stimuli and general lineup performance
